---
title: "Longitudinal Analysis"
author: "Kennedy Zinn"
date: "2025-07-21"
output: html_document
---


```{r}
suppressPackageStartupMessages(library(minfi))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(DMRcate))
suppressPackageStartupMessages(library(maxprobes))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(missMethyl))
suppressPackageStartupMessages(library(MethylToSNP))
```

```{r}
library(lme4)
```

# Load data
```{r}
# load the idat files
rgSet <- read.metharray.exp("/Users/kennedyzinn/Desktop/OneDrive - UTHealth Houston/GRA/TS/dna methyl/SPMTS2_Array data/", force = T)
```

```{r}
# load sample sheet
sample_sheet <- read_excel("/Users/kennedyzinn/Desktop/OneDrive - UTHealth Houston/GRA/TS/dna methyl/Batch3_Sample Sheet_modified.xlsx",
                         skip = 9,
                         na="NA",
                         n_max=8
                         )
sample_sheet <- sample_sheet[,1:7] %>% 
  mutate(lineage = ifelse(Phase!="EC", 0, 1)) %>% 
  dplyr::filter(Karyotype!="46XX")
rownames(sample_sheet) <- paste(sample_sheet$`Sentrix ID`, sample_sheet$`Sentrix Position`, sep = "_")
```

# Pre-processing

```{r}
source("/Users/kennedyzinn/Desktop/Prakash Lab/TS Batch3 DNA Methylation Analysis/process_methyl_microarray.R")
gmSet <- process_methyl_microarray(rgSet)
```

Beta values, which range between [0,1], represent the proportion of methylated DNA copies at a specific CpG site compared to all of the DNA copies at that site. Beta values for the ith CpG site are mathematically defined as,

Beta(i) = max(yi(methyl),0) / [max(yi(unmethyl), 0) + max(yi(methyl), 0) + alpha],

where yi methyl and unmethyl are the intensities measured by the ith methylated and unmethylated probes, respectively. 0 ensures that negative signals that take place after background adjustment are not considered. Alpha, a constant (usually 100), serves as an offset to regularize beta when both methylated and unmethylated instensities are low. Beta values are intuitive, and hence useful for visual presentations of the reads. (Du et al., 2010)

```{r, echo = FALSE}
# Compute beta values
beta_values <- getBeta(gmSet) 
beta_values <- beta_values[,colnames(beta_values) %in% rownames(sample_sheet)]
colnames(beta_values) <- sample_sheet$`Sample Name`
```

M values are log transformed beta values. Negative M values indicate higher levels of unmethylated DNA copies than methylated DNA copies. Postive M values indicate higher levels of methylated DNA copies than unmethylated. M values close to zero indicate similar intensities between methylated and unmethylated probes. M values are mathematically expressed as,

M(i) = log2([max(yi(methyl))+alpha]/[max(yi(unmethyl))+alpha])

where alpha is a constant to offset dramatic changes resulting from small estimation errors. M values can be expressed in terms of beta values,

M(i) = log2(Beta(i)/[1 - Beta(i)])

and therefore, beta values can be expressed in terms of M values,

Beta(i) = 2\^M(i) / [2\^M(i) + 1]

M values are more statistically sound than beta values.

```{r}
# Compute M values
Mval <- getM(gmSet)
Mval <- Mval[,colnames(Mval) %in% rownames(sample_sheet)]
dim(Mval)
```


```{r}
# PCA of M values
par(mfrow=c(1,1), cex =.5)
plotMDS(Mval, labels=sample_sheet$Karyotype, col=as.integer(factor(sample_sheet$Phase)))
legend("top",legend=c("IPSC", "NCC", "EC"), col=as.integer(factor(sample_sheet$Phase)))
```

PC1 and PC2 by disease status

```{r}
par(mfrow=c(1,1), cex =.5)
plotMDS(Mval, labels=sample_sheet$`Sample Name`, col=as.integer(factor(sample_sheet$Karyotype)))
legend("top",legend=c("46XX", "45X", "46XisoXq"), col=1:2)
```

# Differential Methylation Analysis

```{r}
# match column names of Mval to row names of sample sheet
sample_sheet <- as.data.frame(sample_sheet)
rownames(sample_sheet) <- paste(sample_sheet$`Sentrix ID`, sample_sheet$`Sentrix Position`, sep = "_")

# reorder sample sheet to match row names to column names of Mval
sample_sheet <- sample_sheet[colnames(Mval),]

# regular interaction model
sample_sheet <- sample_sheet %>% 
  dplyr::mutate(kary_f = ifelse(Karyotype!="45X", 0,1))

disease <- factor(sample_sheet$kary_f)
lineage <- factor(sample_sheet$lineage)
```

```{r}
design <- model.matrix(~lineage*disease)

fit <- lmFit(Mval, design)
fit2 <- eBayes(fit)
top <- topTable(fit2, number = "Inf")
top
```
```{r}
# difference in methylation changes from ipsc OR NCC to EC in 45x compared to 46XisoXq
topTable(fit2, coef = 4)
```

## Kegg Region Analysis
```{r}
myAnnotation <- cpg.annotate(datatype = "array", object = Mval, what = "M",
                             arraytype = "EPICv1",
                             analysis.type = "differential", design = design, contrasts = F, coef = 4)
DMRs <- suppressMessages(dmrcate(myAnnotation))
results.ranges <- extractRanges(DMRs, genome="hg38")
```

```{r}
library(IlluminaHumanMethylationEPICanno.ilm10b5.hg38)
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b5.hg38)

region <- goregion(results.ranges, all.cpg = rownames(top), collection = "KEGG", array.type="EPIC", anno=anno, sig.genes=T)
```
```{r}
topGSA(region, number = "Inf")
```

```{r}
# pull DMRs mapped to x chromosome
x.ranges <- results.ranges[seqnames(results.ranges)=="chrX",]
x_gsa <- goregion(x.ranges, all.cpg = rownames(top), collection = "GO", array.type="EPIC", anno=anno, sig.genes=T)
topGSA(x_gsa, number = "Inf")
```
```{r}
# pull list of significant genes
library(tibble)
library(tidyr)
x_results <- topGSA(x_gsa, number = "Inf")
x_genes <- x_results%>% 
  dplyr::mutate(SigGenes = strsplit(as.character(SigGenesInSet), split=",")) %>% 
  unnest(SigGenes)
gene_list <- unique(x_genes$SigGenes)
```

```{r}
write.csv(gene_list, "x_genes_from_interaction.csv")
```


```{r}
# pull cpgs associated with significant genes
library(tibble)
library(tidyr)
anno.df <- as.data.frame(anno)
anno.df <- rownames_to_column(anno.df, var = "ID")

top <- topTable(fit2, number = "Inf", coef = 4) %>%
  rownames_to_column(var = "ID")
top <- merge(top, anno.df, by = "ID") %>% # split gene list into individual genes
  dplyr::mutate(UCSC_RefGene_Name = strsplit(as.character(UCSC_RefGene_Name), split = "[,;]")) %>%
  unnest(UCSC_RefGene_Name)
top1 <- top[top$UCSC_RefGene_Name==gene,]
```

```{r}
#average methylation status change difference (compared to control)
library(glue)
for (i in 1:4){
  avg_logfc <- round(mean(results$result[[i]]$logFC), 3)
  print(glue("The average LogFC for {results$gene[i]} in {results$coef[i]} is {avg_logfc}"))
}
```
# Visualize profiles for each sig gene

As only OTX1 actually overlapped with the DMRs, probes used for methylation level were those annotated to the gene instead.
```{r}
# plot beta value per ID gene x group x phase
library(GenomicRanges)
library(ggplot2)

genes <- c("OTX1", "SDK1", "BLOC1S2")

for (gene in genes){
  
  cpgs_on_gene <- results[results$coef=="ipsc_to_smc"&results$gene==gene,]$result[[1]]$ID

  gene_beta <- beta_values[cpgs_on_gene,]

  # Convert matrix to long format
  gene_long <- as.data.frame(gene_beta) %>%
    rownames_to_column("probe_id") %>%
    pivot_longer(-probe_id, names_to = "Sample Name", values_to = "beta") %>%
    left_join(sample_sheet, by = "Sample Name") # merge subject info with beta values

  # Compute average methylation level of gene for each subject and average for each group
  gene_avg <- gene_long %>%
  group_by(ID, Status, Phase) %>%
  summarise(mean_beta = mean(beta, na.rm = TRUE), .groups = "drop")

  group_avg <- gene_avg %>%
  group_by(Status, Phase) %>%
  summarise(mean_beta = mean(mean_beta, na.rm = TRUE), .groups = "drop")

  # plot
  profile <- ggplot() +
    # Individual subject lines
    geom_line(data = gene_avg, aes(x = Phase, y = mean_beta, group = ID, color = Status), linewidth = 1) +
    
    # Group mean dashed lines
    geom_line(data = group_avg, aes(x = Phase, y = mean_beta, group = Status, color = Status), linetype = "dashed", linewidth = 2) +
    scale_color_manual(values = c("control" = "lightblue", "patient" = "red")) +
    labs(title = glue("Methylation Profile of {gene}"), x = "Phase", y = "Methylation Level (β-values)", color = "Phase") +
    theme_minimal(base_size = 14) +
    geom_text(data = gene_avg[gene_avg$Phase=="SMC",], aes(x = Phase, y = mean_beta, label = ID, color = "black"),
              hjust = -0.1, vjust = 0.5, size = 4, show.legend = FALSE)
  print(profile)
}
```


