---
title: "DNA methylation analysis of SMC phase"
author: "Kennedy Zinn"
date: "2025-02-07"
output: html_document
toc: TRUE
---

```{r, echo = FALSE}
suppressPackageStartupMessages(library(minfi))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(DMRcate))
suppressPackageStartupMessages(library(maxprobes))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(missMethyl))
suppressPackageStartupMessages(library(MethylToSNP))
library(RColorBrewer)
```

# Load data
```{r}
# load the idat files
rgSet <- read.metharray.exp("/Users/kennedyzinn/Desktop/OneDrive - UTHealth Houston/GRA/TS/dna methyl/SPMTS2_Array data/", force = T)
```

```{r}
# load sample sheet
sample_sheet <- read_excel("/Users/kennedyzinn/Desktop/OneDrive - UTHealth Houston/GRA/TS/dna methyl/Batch3_Sample Sheet_modified.xlsx",
                         skip = 9,
                         na="NA",
                         n_max=8
                         )
sample_sheet <- sample_sheet[,1:7] %>% 
  dplyr::filter(Phase!="NCC")
rownames(sample_sheet) <- paste(sample_sheet$`Sentrix ID`, sample_sheet$`Sentrix Position`, sep = "_")
```

# Pre-processing

```{r}
source("/Users/kennedyzinn/Desktop/Prakash Lab/TS Batch3 DNA Methylation Analysis/process_methyl_microarray.R")
gmSet <- process_methyl_microarray(rgSet)
```

Beta values, which range between [0,1], represent the proportion of methylated DNA copies at a specific CpG site compared to all of the DNA copies at that site. Beta values for the ith CpG site are mathematically defined as,

Beta(i) = max(yi(methyl),0) / [max(yi(unmethyl), 0) + max(yi(methyl), 0) + alpha],

where yi methyl and unmethyl are the intensities measured by the ith methylated and unmethylated probes, respectively. 0 ensures that negative signals that take place after background adjustment are not considered. Alpha, a constant (usually 100), serves as an offset to regularize beta when both methylated and unmethylated instensities are low. Beta values are intuitive, and hence useful for visual presentations of the reads. (Du et al., 2010)

```{r, echo = FALSE}
# Compute beta values
beta_values <- getBeta(gmSet) 
beta_values <- beta_values[,colnames(beta_values) %in% rownames(sample_sheet)]
colnames(beta_values) <- sample_sheet$`Sample Name`
```

M values are log transformed beta values. Negative M values indicate higher levels of unmethylated DNA copies than methylated DNA copies. Postive M values indicate higher levels of methylated DNA copies than unmethylated. M values close to zero indicate similar intensities between methylated and unmethylated probes. M values are mathematically expressed as,

M(i) = log2([max(yi(methyl))+alpha]/[max(yi(unmethyl))+alpha])

where alpha is a constant to offset dramatic changes resulting from small estimation errors. M values can be expressed in terms of beta values,

M(i) = log2(Beta(i)/[1 - Beta(i)])

and therefore, beta values can be expressed in terms of M values,

Beta(i) = 2\^M(i) / [2\^M(i) + 1]

M values are more statistically sound than beta values.

```{r}
# Compute M values
Mval <- getM(gmSet)
Mval <- Mval[,colnames(Mval) %in% rownames(sample_sheet)]
dim(Mval)
```

```{r}
# sanity check
par(mfrow=c(1,1), cex =.5)
plotMDS(Mval, labels=sample_sheet$`Sample Name`, col=as.integer(factor(sample_sheet$Karyotype)))
legend("top",legend=c("46XX", "45X", "46XisoXq"), col=1:2)
```

## Differential Methylation Analysis

# Differential Methylation Analysis

```{r}
# match column names of Mval to row names of sample sheet
sample_sheet <- as.data.frame(sample_sheet)
rownames(sample_sheet) <- paste(sample_sheet$`Sentrix ID`, sample_sheet$`Sentrix Position`, sep = "_")

# reorder sample sheet to match row names to column names of Mval
sample_sheet <- sample_sheet[colnames(Mval),]

# regular interaction model
phase <- factor(sample_sheet$Phase, labels=c("iPSC", "EC"))
```

```{r}
# Phase
design <- model.matrix(~phase, data = sample_sheet)
fit <- lmFit(Mval, design)
fit2 <- eBayes(fit)
top <- topTable(fit2, number = "Inf")
top
```


## KEGG Region Level Analysis

```{r}
myAnnotation <- cpg.annotate(datatype = "array", object = Mval, what = "M",
                             arraytype = "EPICv1",
                             analysis.type = "differential", design = design, contrasts = F, coef=2)
DMRs <- suppressMessages(dmrcate(myAnnotation))
results.ranges <- extractRanges(DMRs, genome="hg38")
```

```{r}
library(IlluminaHumanMethylationEPICanno.ilm10b5.hg38)
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b5.hg38)

region <- goregion(results.ranges, all.cpg = rownames(top), collection = "KEGG", array.type="EPIC", anno=anno, sig.genes=T)
```

```{r}
topGSA(region, number = "Inf")
```

```{r}
# pull DMRs mapped to x chromosome
x.ranges <- results.ranges[seqnames(results.ranges)=="chrX",]
x_gsa <- goregion(x.ranges, all.cpg = rownames(top), collection = "GO", array.type="EPIC", anno=anno, sig.genes=T)
topGSA(x_gsa, number = "Inf")
```
```{r}
# set up the grouping variables and colors
cols <- c(2,4)[phase]
names(cols) <- phase

# draw the plot for the top DMR
par(mfrow=c(1,1))
DMR.plot(ranges=x.ranges, dmr=1, CpGs=beta_values, phen.col=cols, what = "Beta",
         arraytype = "EPICv1",
         genome="hg38")
```

# Pathway Analysis by Methylation Direction

## Identify the direction of significant gene sets

```{r}
top <- topGSA(x_gsa, number = "Inf")
sig.GeneSets <- unique(top$SigGenesInSet)
# Split the gene sets into individual genes
split_genes <- function(gene_set) {
 unlist(strsplit(gene_set, split = "[,]"))
}

# Apply the splitting function to all significant gene sets
all_genes <- unique(unlist(lapply(sig.GeneSets, split_genes)))
```

```{r}

# Annotate probes
library(tibble)
top <- rownames_to_column(top, var = "ID")
anno <- as.data.frame(anno)
anno <- rownames_to_column(anno, var = "ID")
top <- merge(top, anno, by = "ID")
```

```{r}
# Split UCSC_RefGene_Name into individual genes
library(tidyr)
top <- top %>%
  mutate(UCSC_RefGene_Name = strsplit(as.character(UCSC_RefGene_Name), split = "[,;]")) %>%
  unnest(UCSC_RefGene_Name)

# Filter for significant gene sets
filteredResults <- top[top$UCSC_RefGene_Name %in% all_genes, ]
```

```{r}
# Summarize methylation status for each significant gene set
methylationSummary <- sapply(all_genes, function(gene) {
 probes <- filteredResults[filteredResults$UCSC_RefGene_Name == gene, ]
 avgLogFC <- mean(probes$logFC)
 return(avgLogFC)
})
```

```{r}
# Determine methylation status
methylationStatus <- ifelse(methylationSummary > 0, "Hypermethylated", "Hypomethylated")

# Combine results into a data frame
results <- data.frame(GeneSet = all_genes, MethylationStatus = methylationStatus)

print(results)
```

## Pathway analysis of genes by methylation direction

```{r, echo = F}
library(clusterProfiler)
library(org.Hs.eg.db)
source("/Users/kennedyzinn/Desktop/OneDrive - UTHealth Houston/GRA/dna methyl/gene.list.R")
# create gene list of hypermethylated genes
hyper.gene.list <- gene.list("Hypermethylated")

# create gene list of hypomethylated genes
hypo.gene.list <- gene.list("Hypomethylated")
```

### GO

```{r}
# hypermethylated gene sets
go.hyper.enrich <- enrichGO(gene = hyper.gene.list, OrgDb = org.Hs.eg.db, keyType = "ENTREZID", ont = "ALL")

# hypomethylated gene sets
go.hypo.enrich <- enrichGO(gene = hypo.gene.list, OrgDb = org.Hs.eg.db, keyType = "ENTREZID", ont = "ALL")

# add column specifying methylation direction
go.hyper.enrich <- as.data.frame(go.hyper.enrich)
go.hyper.enrich$MethylationStatus = "Hypermethylated"

go.hypo.enrich <- as.data.frame(go.hypo.enrich)
go.hypo.enrich$MethylationStatus = "Hypomethylated"

# merge data sets
go.enrich <- rbind(go.hyper.enrich, go.hypo.enrich)
```

### KEGG

```{r}
# hypermethylated gene sets
kegg.hyper.enrich <- enrichKEGG(gene = hyper.gene.list, organism = "hsa")

# hypomethylated gene sets
kegg.hypo.enrich <- enrichKEGG(gene = hypo.gene.list, organism = "hsa")

# add column specifying methylation direction
kegg.hyper.enrich <- as.data.frame(kegg.hyper.enrich)
kegg.hyper.enrich$MethylationStatus = "Hypermethylated"

kegg.hypo.enrich <- as.data.frame(kegg.hypo.enrich)
kegg.hypo.enrich$MethylationStatus = "Hypomethylated"

# merge data sets
kegg.enrich <- rbind(kegg.hyper.enrich, kegg.hypo.enrich)
```

Certain interesting gene sets can be pulled in order to determine whether the gene set is significantly hyper or hypomethylated.

```{r}
kegg.enrich[kegg.enrich$ID == "hsa04740", 15]
```

# Export Results

```{r}
write.csv(kegg.enrich, "kegg_SMC_enrichment_results.csv")
write.csv(go.enrich, "go_SMC_enrichment_results.csv")

# create csv of gene list
capture.output(compact(hyper.gene.list), file = "SMC_hypermethyl_genes.txt")
capture.output(compact(hypo.gene.list), file = "SMC_hypomethyl_genes.txt")
```
